---
title: "Data Processes Final Project - Group 1"
author: Artjom Poljakow, Wladyslaw Eysymontt, Mateusz Klimas, Juan Luis Ruiz-Tagle,
  Jorge Martín Lasaosa
date: "20/12/2019"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(plotly)

source("Project.R")
```

## Final Project Rubric
This rubric is intended to help guide expectations and increase transparency. However, it is not necessarily fully exhaustive. Make sure to read the detailed explanation on Moodle.

### Abstract (**5 points**)
- Succinctly summarizes the importance and findings of the project within the 150 word limit.

  The intention of this paper is to find a way to influence the incident rate related to guns. As the most amount of data 
  considering this topic is available from the united stated it was the chosen are to work on. Datasets considered were the   
  following, one with incidents happened from 2013-18, evolution of laws considering gun regulation, gun license
  aquirement and population disctribution in the states. 
  To gain information the incidents were related to an amount of people to normalize the value. 
  An metric which shows how many laws are applied to gun regulation in a state and their development over the years is       introduced and explained. 
  The visualized graphs datasets and the applied model show a clear relationship between the following 
     TODO 
  Considering the relationship is can or cannot be advised to apply more laws, change regulation considering the   
     aquirement of guns and a govermental controlled instance which intervend by people who clearly are a threat for the 
     community before incidents happen.
  

### Introduction and Related Work (**10 points**)
- Provides a clear motivation for answering a _specific_ data driven question of interest (**5 points**)

  As the clear problem of weapon abuse among civilians is a public threat to everybody especially among the most volnerable
  the children the motivation is to find a way to influence this behavior in a positiv way, that is to decrease the amount
  incidents happening every year. As there are several ways of handling this problem by taking a look among different 
  countries, there is none with such a broad and detailed amount of data as the united states. As every state provides
  their own laws on this topic, a major part of citizens rely on the constitution. To keep in mind that it is not quite 
  reasonable to collect all weapons the next day and solve the problem this way it is feasable to concentrate on rather
  credible possibilities. Which leads us to our question, do the restrictions in a concrete State in order to purchase or
  use a gun affect the "gun kills/population" rate? While investigating the data we sure took a look on related
  publications as the following.
  

- Cites 5 _relevant_ pieces of relevant work (whatever format you choose is fine, including just a hyperlink) (**1 point each**)
  Under the cirsumstances explained it is quite a harsh decision to make for everybody involved. During the last years
  different states have adopted so called a red flag law. This law enables the court to collect weapons from specific 
  individuals if a threat for them or the community is given. Therefore the judge handling the case need to decide if
  the evidence, which are the most threats, are sufficient to claim a persons opinion as a threat and restrict their rights
  of possessing a gun and use it in which ever possible way they like. Articles like
  [judge decisions](https://www.nytimes.com/2019/11/18/us/gun-seizures.html), 
  [red flag law](https://www.nytimes.com/2019/08/06/us/red-flag-laws.html), 
  [protection](https://www.econlib.org/the-purpose-of-a-gun-is-not-to-kill/), 
  [suicide](https://lawcenter.giffords.org/facts/gun-violence-statistics/),
  [suicide](https://lawcenter.giffords.org/facts/gun-violence-statistics/),
  [domestic violence](https://lawcenter.giffords.org/gun-laws/policy-areas/who-can-have-a-gun/domestic-violence-firearms/)
  illustrate the situation.
  To relief judges and the cociety the laws have to be adjusted so people do not feel like their rights were restricted 
  and a abuse of firearms is prevented.
  
### Exploratory Data Analysis (**20 points**)

Introduces the dataset by describing the origin (source) and structure (shape, relevant features) of the data being used (**5 points**)

1. [Federal Firearm Licences](https://www.kaggle.com/doj/federal-firearm-licensees) The original data is published online in a tab-separated format by the [Department of Justice Bureau of Alcohol, Tobacco, Firearms, and Explosives](https://www.atf.gov/resource-center/data-statistics). This association, also known as ATF, compiles a comprehensive collection of ATF-related data from national surveys, state-based surveys, other collected license statistics, and other data sources documents trends in firearms, commerce and use of federal services in the United States.There are important features such as:

    * State to which belongs the license
    * City to which belongs the license
    * Type of license


2. [Gun violence Dataset](https://www.kaggle.com/jameslko/gun-violence-data): More than 260.000 gun violence incidents in the US between January 2013 and March 2018 with detailed information about each one. GVA organization utilize automated queries, manual research through over 6,500 sources from local and state police, media, data aggregates, government and other sources daily. Each incident is verified by both initial researchers and secondary validation processes. There are relevant information such us:

    * the location of the incident
    * participants
    * date
    * source of information
    * incident id


The main structure is **blablablablabla**. However, the combination of categorical and continuous features has plenty of null values. This can be harmful to the machine learning algorithms. 

3. [Firearms Provisions in US States](https://www.kaggle.com/jboysen/state-firearms): The State Firearm Laws project aims to provide researchers with the data necessary to evaluate the effectiveness of various firearm laws. This dataset covers all 50 US States from 1991 to 2017 and includes 133 binary variables which explain the states requirements. These requirements are related to the following topics:
    
    * Vendor license required to sell ammunition.
    * Records of ammunition sales must be retained by the dealer.
    * Permit required to purchase ammunition.
    * Background checks required for ammunition purchases.
    * Sale of ammunition is restricted to the same categories of those who are legally allowed to purchase firearms.
    * Purchase of any type of ammunition restricted to those ages 18 and older.
    * Purchase of handgun ammunition restricted to those ages 21 and older.

    One-hundred of the 133 provisions were coded by Michael Siegel, MD, MPH, Boston University School of Public Health, with funding from the Robert Wood Johnson Foundation, Evidence for Action: Investigator-Initiated Research to Build a Culture of Health program (grant #73337), using data derived from the Thomson Reuters Westlaw legislative database. The other 33 provisions were coded using a database created by Everytown for Gun Safety and Legal Science, LLC. Shared in accordance with the Creative Commons Attribution-4.0 International License.

4. [US Mass Shootings, 1982-2019](https://www.motherjones.com/politics/2012/12/mass-shootings-mother-jones-full-data/):
This dataset is based on information recompiled from several public sources, such as Wikipedia, Mother Jones, Stanford, USA Today and others. It contains only 323 rows distributed across 50 years and 49 states; therefore, the data is very dispersed and may not be useful for predictions. Even though, it can help us to understand some general behaviours and relations between the variables, especially since it is very strongly related to our other datasets. Different data transformations, such as unification of State’s names and deleting rows containing no State name had been done for being able to continue with an explanatory analysis.
The graphic made on its data, “Victims of mass shootings per State” is a stacked-bar plot which represents for every State the sum of the people injured and killed during all the years. It can help us to understand not only what States suffered more from these events, but also where the shooting are usually more deadly by the rate of injured/killed people. Some important variables are:
  * total_victims
  * weapons_obtained_legally
  * weapon_type

#### Results of data set exploration

##### 1. Heatmap graph representing "Firearms Provisions in US States" dataset
    
    For each state and year, the percentage of policy measures are represented. Although a lot of states are improving their provisions, the mayority do not reach even a half of them. WRITE MORE.
    ```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
    plot_1_heatmap
    ```
##### 2. Evolution of injured and killed victims in USA between 2013 and 2018
    
    This graph showcases the evolution of the number of people killed and injured  in USA from 2013 to 2018. We notice quickly that the data from 2013 is far from complete, even though they have entries for all the months. 2018 has normal data from January to March, but then it finishes abruptly. We can see how the number of injured people doubles the number of killed people throughout the whole time period. If we watch closely we can perceive a tendency, some sort of cycle that repeats every year where the number of injured and killed people decreases during the winter, while it has peaks on summer.
    
    ```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}

    ggplotly(plot_2_line)
    
    ```
##### 3. HeatMap with gun selling licenses in the USA
    First, in order to design the heatmap, we have merged the columns Premise.State and Mail.State into one, checking if both of them are the same in order to don't assume the place where this license belong, as the publisher doesn't specify to what these two different columns of states refer to. Also, the columns like state are converted from factors to string. 
  To get the data ready to use for the USA map, we first group the observations by state and license type. Then we add all the licenses to get only groups for each type of license with the number of licenses, which later will be added together in order to get the number of licenses for each state.
To create the heatmap, we use the library usmap, using the function plot_usmap, to which me input the data we have processed, and where we define things like the low and high colors of the heatmap, the legends, etc. 
Once the heatmap is ploted, we can observe all the states with a different color grade, where Texas is the state with more licenses. The other states display different ranges of colours.
    
    ```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}

    ggplotly(plot_3_licenses_map)

    ```

##### 4. Barchart with victims of mass shootings per state.
    
    Description
    
    ```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}

    ggplotly(plot_4_victims_state)

    ```



##### 5. Map with incidents by location.
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
    complete_Map

```
   
   
### Methods (**30 points**)
The question of interest selected from the proposal is the next one:

<center> _Do the restrictions in a concrete State in order to purchase or use a gun affect the "gun kills/population" rate?_ 
</center>

In order to answer the question, "Firearms Provisions in US States" and "Gun violence" datasets were joined together. By doing this we have been able to understand the impact of the restrictions in each specific state on the number of killed victims throughout the years. We have created two new variables:
- Score: Score is just the percentage of provisions (policies) that were active in a given state at a given year. 
- Rate: Rate is the ratio of killed victims per state population in a year.

Our hypothesis states that there is a significant negative correlation between these two variables. We believe that *Year* has also impact on the final rate (we saw in graph 2 that there is a growing tendence of criminal activity), and that *State* has a huge dependence on the *Rate* since it is a well known fact that some states are way more violent than others (just have a glance at graph 3).

This hypothesis can be represented by the following formula:

$$Rate = \beta_0 + \beta_1 Score + \beta_2 State + \beta_3Year$$


To confirm or discard this hypothesis we will start by looking at the correlation matrix between these variables. 

```{r echo=FALSE, fig.align='center'}
    
    #ggcorrplot(correlation_matrix_complete_plot, lab = T, title = "Corr matrix on selected variables")
    ggcorrplot(correlation_matrix_complete_plot, lab = T, title = "Corr matrix on selected variables")
    
    ```
We can see how there is a strong inverse correlation between *Score* and our target variable *Rate*. As well ther is a mild correlation between *Year* and *Rate* and absolutely none (0.01) between *Year* and *Score*. This is exactly what we need to get a good linear regression: correlation between the variables and the target and independence between the variables themselves. If this is the case we are sure to avoid multicollinearity issues. There are other assumptions that should be checked (homeostaticity, normality of residuals, etc) but these lie beyond the scope of this project. The *State* variable is not included in the matrix since it is a categorical variable.

Taking all this into consideration, we proceed to create our model in R. To do this we first check that there are no missing values in the dataset. In order to use the *State* as a categorical variable we make sure that it is configured as a `factor`, which it is by default. Then we split the dataset in training and test data in a 75% - 25% ratio respectively. Afterwards we start training our model using cross validation with 10 folds and 3 repetitions and gridsearch. The only parameter our grid will search is the intercept (the only one available for linear regression).

### Results (**20 points**)

The results that we obtain from our training model are the following.
    ```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
    #linear_model
    ```

Regarding the strength of the variables, the upcoming graph plots how decisive each variable is in the overall prediction of the model.

    ```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}

    ggplotly(varimp_plot)
    
    ```

We see in this graph how some of the states are critical to the estimation of the *Rate* response variable.



    ```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
  
  ##This is for whatever reason commented on the html output, thererfore commented here TO SOLVE
  est_rate <- ggplot(test, aes(rate,rate_prediction)) +
  geom_point() +
  geom_smooth(method = "lm", se = T)
  #  ggtitle("Relation between predicted and target variables")
    
  ggplotly(est_rate)

    ```

We have seen that the performance measures are very excellent on this model (RSME, Rsquared and  MAE are very low). The graph above compares the predicted values with the targets, and we notice how the proportion is almost 1:1. 

### Discussion and Future Work (**10 points**)
Based on _specific observations_ from the results section, the report clearly provides:
  - An analysis of the real world implications of the results, at least one full paragraph (**5 points**)
  
  Based on specific observations from the results section, the report clearly provides:

    An analysis of the real world implications of the results, at least one full paragraph (5 points)
	
	Based on the information gained through the analysis, the hypothesis that a restriction by law to get guns will reduce the
	amount of incidents, can be confirmed. The amount of incidents in states where the purchase of a gun was restrictive
	or got more restrictive by the years did drop. Nevertheless the fact that a high amount of citizens implies a appropriate
	amount of incidents should not be ignored. Therefore an analysis of the amount of citizens was taken in account. Recpectively
	the states with the highes amount of incidents with refering to their inhabitants are Alaska, Delaware, Louisiana, 
	South Carolina and Illinois. 
	Considering the score created to rank the states by their application of law considering gun restrictions the states mentioned
	obtain a score below 20 throught the years from 1991 till 2017. This includes Alaska 10 -> 5, Louisiana ~10, 
	South Carolina 15 -> 10, 
	slight improvement in Delaware, on the contrary Illinois, despite having higher score compared to the mentioned before, still
	has a high rate of incidents.(30 -> 50)
	Now taking a look at the states with the lowest amount of incidents are Hawaii, Arizona, Utah, Idaho, California. They all
	have an incident rate below 50 per 100k of inhabitants.  
	Hawaii(30 -> 60), 
	Arizona(10), 
	Utah(10), 
	Idaho(5), 
	California(50 -> 80)
	That questions the created score ranking. Nevertheless the score is created equaly across all laws available,
	this implies that all laws have an equal impact on the costumer, which it certainly has not. To improve the score 
	and create a more reliable metric a weighted sum can be considered.
  
  - Clear suggestion for directions of future research, at least one full paragraph (**5 points**)

### Code Quality (**5 points**)
Code is well commented and structured (e.g., indented), organized across multiple different files, uses clear variable names, and runs on any computer.
